---
title: "SAM Diagnostics for the Jack Mackerel Assessment"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: "Niels T. Hintzen"
format:
  html:
    code-fold: true
    code-summary: "Show code"
    fig-cap-location: bottom
    number-sections: true
    theme: readable
    toc: true
    toc-depth: 3
    toc-location: right
    lightbox: true
bibliography: refs.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE, message = FALSE, echo = FALSE,
  fig.width = 9, fig.height = 6
)

diag_dir <- normalizePath(
  file.path(dirname(knitr::current_input(dir = TRUE)), "diagnostics"),
  winslash = "/",
  mustWork = FALSE
)

required_diag <- file.path(diag_dir, "h1_stock_trajectory.png")
if (!file.exists(required_diag)) {
  if (!file.exists("output_h1.RData")) {
    stop(
      "Missing diagnostics/ outputs and output_h1.RData. ",
      "Run 01_run_assessment.R first or provide output_h1.RData."
    )
  }
  source("generate_diagnostics_assets.R")
}

show_diag <- function(file, caption = NULL, width = "100%") {
  path <- file.path(diag_dir, file)
  knitr::include_graphics(path)
}
```

# Introduction {#sec-intro}

## Background and motivation

The Chilean jack mackerel (*Trachurus murphyi*) stock in the South Pacific is assessed
annually by the SPRFMO Scientific Committee using the Joint Jack Mackerel Model (JJM),
an age-structured statistical catch-at-age model implemented in AD Model Builder
[@fournier2012ad]. Two competing stock-structure hypotheses are evaluated in parallel:
hypothesis H1 (a single shared stock ranging from Peru to the offshore zone) and
hypothesis H2 (a separate far-northern stock off Peru combined with a central-southern
stock off Chile). The SC13 base model (`h1_1.14` and `h2_1.14`) constitutes the
assessment configuration on which the current MSE programme is being conditioned.

### The SPRFMO Jack Mackerel MSE programme

A Management Strategy Evaluation (MSE) process for CJM has been underway within
SPRFMO for several years and entered a critical delivery phase in 2026. The JM MSE
Task Team (TT), chaired by J. Ianelli, is working through a structured programme
following the Carruthers et al. (2024) MSE roadmap framework. 

A central challenge identified in BM-01 is that the SC13 JJM model is **difficult to
transfer directly into an MSE Operating Model** for four reasons:

1. **Time-varying selectivity** — JJM estimates annual selectivity within blocks, but
   the block structure and its projection behaviour must be simplified and fixed for OM
   conditioning;
2. **Dynamic reference points** — JJM's $B_\text{MSY}$ and $F_\text{MSY}$ are
   re-derived each year and cannot be directly carried into a forward-simulation OM
   without specifying which biological period they represent;
3. **Differing fitting and projection assumptions** — the estimation model and the
   projection model in JJM use different assumptions (e.g. about recruitment) that must
   be reconciled before conditioning;
4. **Mixed stock-structure differences** — H1 and H2 require consistent treatment of
   the far-northern component to allow meaningful comparison across hypotheses.

BM-01 proposes four simplified alternatives to address these issues: (§6.1) block-based
selectivity, (§6.2) a single S–R relationship per hypothesis without regime splitting,
(§6.3) fixed static reference points, and (§6.4) harmonised northern treatment across
H1 and H2. The TT is scheduled to **agree the selectivity block structure and the
reference point approach at TT-03 (30 April 2026)**, with the selectivity question
narrowed to at most two candidates at TT-02 (9 April 2026).

For the stock-recruitment relationship (BM-04), the current working prior for H1 is
steepness $h = 0.65$ with recruitment variability $\sigma_R = 0.6$. An ENSO
offshore-drop scenario is also planned as a key robustness test. The OM will be
implemented and validated across three platforms — FLR, RTMB, and OpenMSE (DLMtool /
MSEtool) — with cross-platform validation addressed in JM-WP-SC14-02.

### Role of this SAM analysis

The SAM model is applied here to the H1 configuration as an **independent diagnostic
tool**, not as a replacement for JJM. Its state-space structure and internal
data-weighting make it particularly well suited to addressing the open OM-design
questions described above — questions that are difficult to resolve from within the
JJM framework alone:

- **Selectivity** (@sec-blocks): SAM's annual random-walk F provides an objective,
  statistically grounded basis for choosing how many selectivity blocks are needed and
  where their boundaries should fall, weighted by the catch volume in each period.
- **Productivity and reference points** (@sec-productivity): The period-specific
  Beverton-Holt steepness estimated here can be compared against the working prior
  $h = 0.65$ and used to compute block-specific $F_\text{MSY}$ for the OM.
- **Index uncertainty and fleet conflict** (@sec-q, @sec-conflict): SAM's internally
  estimated observation variances and the cross-fleet residual diagnostics identify
  which of the seven abundance indices carry the most information, which are
  redundant, and which show systematic trends — directly informing BM-03 (*Indices of
  Abundance: Review and Standardisation*) and the choice of MP input data streams
  for BM-04.

## The State-Space Assessment Model (SAM)

### Overview and ICES context

SAM was developed at the Technical University of Denmark and is maintained as an open
R package (`stockassessment`). It has become the **standard stock assessment model for
the majority of ICES-advised stocks**, including North Sea herring, North Sea cod,
mackerel, blue whiting, and many others. Although it is not yet widely used outside
the North Atlantic, its mathematical foundations are general and apply directly to
multi-fleet pelagic assessments such as CJM [@nielsen2021multifleet].

The core idea is to treat the unobserved true population state — numbers-at-age and
fishing mortality — as **latent random variables** that evolve through time according
to a process model, while catch and survey observations are linked to those states
through an observation model with estimated error variances. This formulation is known
as a **state-space model** and is fitted using the **Laplace approximation**
implemented in the **TMB** (Template Model Builder) automatic differentiation
framework [@kristensen2016tmb]. TMB integrates out the latent states numerically,
making the Laplace approximation necessary because the catch equation's Baranov term
renders the model nonlinear in the latent states. The result is a
computationally efficient, fully differentiable likelihood that propagates both process
and observation uncertainty into all derived quantities.

### Population process model

Log-abundance at age follows a random-walk survival equation with process noise:

$$
\log N_{1,t} \sim \mathcal{N}\!\bigl(\mu_R,\,\sigma_R^2\bigr)
\qquad\text{(recruitment, treated as a free random effect)}
$$

$$
\log N_{a,t} = \log N_{a-1,\,t-1} - Z_{a-1,\,t-1} + \varepsilon_{a,t},
\quad \varepsilon_{a,t} \sim \mathcal{N}(0,\sigma_a^2)
$$

The process error $\varepsilon_{a,t}$ absorbs year-to-year deviations from the
deterministic survival equation — deviations that in a classical statistical
catch-at-age model would be forced into the residuals of the catch or survey
likelihood. This means SAM can fit the data more honestly without over-shrinking
recruitment or distorting the F trajectory to compensate for a poor-fitting year.

Fishing mortality also evolves as a latent random walk:

$$
\log F_{a,t} = \log F_{a,t-1} + \eta_{a,t},
\quad \eta_{a,t} \sim \mathcal{N}(0,\sigma_{F,k}^2)
$$

Ages that share the same variance parameter $\sigma_{F,k}^2$ are **coupled**: they
move together through time, effectively defining a selectivity group. The number of
F-variance groups, and which ages belong to each group, is a key user choice (see
@sec-mse-config). Fewer groups impose a smoother, more constrained selectivity; more
groups allow independent trajectories per age.

### Observation model

Catch-at-age observations are log-normal:

$$
\log\hat{C}_{a,t,f} \sim \mathcal{N}\!\bigl(\log C_{a,t,f}^{\text{pred}},\,\sigma_{C,f}^2\bigr)
$$

Survey biomass indices (as used here, since the CJM surveys are treated as integrated
biomass signals) are:

$$
\log\hat{I}_{s,t} \sim \mathcal{N}\!\bigl(\log(q_s\,\hat{B}_{s,t}),\,\sigma_{I,s}^2\bigr)
$$

where $\hat{B}_{s,t}$ is the model-predicted exploitable biomass for survey $s$. The
**observation variances $\sigma^2$ are estimated from the data**, not fixed externally.
This is a fundamental difference from JJM, where data weights are set by the analyst.
SAM's internally estimated variances provide a direct, objective measure of the
effective weight each data source receives — large variance = low effective weight.

### Uncertainty propagation

Because SAM uses the Laplace approximation, the full posterior of all latent states
and derived quantities (SSB, $\bar{F}$, recruitment) is approximated as multivariate
normal. This gives analytical confidence intervals for all time-series outputs without
requiring MCMC sampling, making it practical to produce annual uncertainty estimates
for SSB, F, and recruitment routinely.

## The JJM Model

JJM is a forward-projection statistical catch-at-age model. Numbers-at-age evolve
deterministically as:

$$
N_{a,t} = \begin{cases}
R_t & a = 1 \\
N_{a-1,\,t-1} \exp\!\bigl(-Z_{a-1,\,t-1}\bigr) & 1 < a < A \\
N_{A-1,\,t-1} \exp\!\bigl(-Z_{A-1,\,t-1}\bigr) + N_{A,\,t-1}\exp\!\bigl(-Z_{A,\,t-1}\bigr) & a = A
\end{cases}
$$

where $Z_{a,t} = M_{a,t} + F_{a,t}$. Fishing mortality is decomposed over fleets as:

$$
F_{a,t} = \sum_{f=1}^{F} s_{a,f}\,\Gamma_{f,t}
$$

with $s_{a,f}$ selectivity and $\Gamma_{f,t}$ the fully-selected fishing mortality.
Predicted catch follows the Baranov equation [@hilborn1992quantitative]:

$$
C_{a,t,f} = \frac{s_{a,f}\,\Gamma_{f,t}}{Z_{a,t}}\,N_{a,t}
  \bigl(1 - \exp(-Z_{a,t})\bigr)
$$

Survey indices are modelled as:

$$
I_{s,t} = q_s \sum_a v_{a,s}\,N_{a,t}\,w_{a,t}
$$

Observed compositions (age and length) are fitted with a multinomial likelihood.
Data weights are specified externally (e.g. Francis reweighting, @francis2011data).

## Key differences between JJM and SAM

| Feature | JJM | SAM |
|---|---|---|
| **Framework** | Deterministic forward projection (ADMB) | State-space model with Laplace approximation (TMB) |
| **Process error** | None — $N$ fully determined by catch | Random walk in $\log N$ and $\log F$ |
| **Composition data** | Age and length proportions (multinomial) | Age proportions only (log-normal) |
| **Data weighting** | External (Francis, input ESS) | Internal — $\sigma^2$ estimated |
| **Selectivity** | Constant within user-defined year blocks | Annual random walk; blocks optional |
| **Catchability** | Fixed $q$ per survey | Fixed $q$; residual trends reveal drift |
| **Reference points** | Computed internally per selectivity block via SPR/YPR | Externally estimated |
| **Uncertainty** | Hessian-based delta method | Laplace approximation; MCMC optional |

: Key differences between JJM and SAM {#tbl-compare}

## SAM in the MSE context {#sec-mse-config}

The MSE process for CJM requires an Operating Model (OM) that can generate plausible
stock trajectories under different harvest control rules. The OM is typically
conditioned on the assessment model: its parameter estimates, uncertainty structure,
and dynamic assumptions are inherited directly from the assessment. Three open
questions are particularly important for the OM design, and SAM provides evidence
for each.

### 1. Selectivity: constant blocks versus annual random walk

JJM parameterises selectivity as constant within user-defined year blocks. The OM
inherits these blocks directly, meaning the OM's F trajectory in the projection period
is governed by the selectivity pattern of the most recent block. If that block does not
accurately represent the current fleet behaviour — because it is too long, or because
the true selectivity has been gradually shifting — the OM will be biased.

SAM estimates selectivity annually via a random walk and therefore reveals *when* the
selectivity shape changed significantly. The block analysis in @sec-blocks translates
those SAM-estimated change points into suggested block boundaries for JJM, ranked by
importance. The OM designer can then choose the minimum number of blocks that captures
the dominant selectivity regimes without overfitting.

The key trade-off is:

- **Too few blocks**: selectivity is over-smoothed; the OM misrepresents the current
  fleet's age targeting and will generate biased F projections.
- **Too many blocks**: each block is estimated from fewer data, increasing uncertainty
  in selectivity; the OM carries more estimation noise into projections.

### 2. Biological reference points: fixed versus time-varying

JJM already computes biological reference points ($F_\text{MSY}$, $B_\text{MSY}$,
$B_{40\%}$) internally via equilibrium per-recruit (SPR/YPR) analysis within each
selectivity block. Because JJM estimates a separate selectivity pattern for each
block, the reference points it reports are inherently block-specific: changing the
selectivity block structure directly changes the reference point estimates. This is
the mechanism through which time-varying reference points enter the assessment —
there is no need to compute them externally.

The key question for the OM is therefore not *how* to obtain time-varying reference
points, but *whether* the JJM block-specific values show meaningful variation that
should be preserved in the OM, versus collapsing to a single set. The productivity
analysis in @sec-productivity provides complementary evidence: if the Beverton-Holt
steepness $h$ shows a statistically supported shift across periods, this corroborates
using the block-specific reference points rather than a time-series average.
Conversely, if steepness is stable across the full time series, a single set of
reference points (derived from the most recent or most representative selectivity
block) is sufficient for the OM.

In practice, the OM designer should:

1. Confirm which JJM selectivity blocks are carried forward into the OM (informed by
   @sec-blocks).
2. Use the JJM-reported $F_\text{MSY}$ and $B_\text{MSY}$ values for each retained
   block directly as the block-specific reference points.
3. Cross-check whether the magnitude of variation in those block-specific reference
   points is large relative to their estimation uncertainty — if not, a single set
   is adequate.

### 3. Which data sources carry the most uncertainty?

The OM must replicate the observation process — how future monitoring data will be
generated and how uncertain they will be. The relative magnitude of SAM's internally
estimated observation variances (@sec-obsvars) directly answers this question: data
sources with large $\hat{\sigma}^2$ contribute little to the current assessment and
will contribute little to future assessments under the same design. The runs test
(@sec-runs) identifies sources with systematic, non-random residuals — these are
candidates for structural changes in the OM's observation model (e.g. time-varying
$q$). Together, these diagnostics provide the evidence base for deciding which surveys
to include in the OM's observation model and what variance to assign them.

# Data and Configuration {#sec-data}

## Input data

The SAM fit uses the H1 single-stock FLStock (`h1_1.14`) prepared for the 2025
assessment cycle. There are four catch fleets and seven biomass-only survey indices
(@tbl-surveys).

| Index | Type | SAM treatment |
|---|---|---|
| Chile_AcousCS | Acoustic survey (central-south Chile) | Total biomass index |
| Chile_AcousN | Acoustic survey (north Chile) | Total biomass index |
| Chile_CPUE | Chilean CPUE | Total biomass index |
| DEPM | Daily egg production method (spawning biomass) | SSB index |
| Peru_Acoustic | Acoustic survey (Peru) | Total biomass index |
| Peru_CPUE | Peruvian CPUE | Total biomass index |
| Offshore_CPUE | Offshore pelagic trawl CPUE | Total biomass index |

: Survey indices used in the SAM fit {#tbl-surveys}

All indices are treated as biomass indices (age = "all") using `FLIndex` objects with
`type = "biomass"`. Acoustic surveys and CPUE indices (Chile, Peru, Offshore) are all
fitted as total biomass indices (`biomassTreat = 5`); DEPM is fitted as an SSB index
as it directly estimates egg production from spawning adults.

## SAM control configuration

Four catch fleets share age-coupled N states (linked cohorts) while F states are estimated
independently for each fleet. Within each catch fleet, selectivity-at-age is controlled by
shared N-state groups (e.g. ages 1–7 linked for N_Chile, ages 1–8 for SC_Chile_PS).
F-variance and observation-variance parameters for the catch fleets use a single shared
parameter per fleet. Surveys each receive an
independent observation-variance parameter and are linked to total
biomass except DEPM which uses SSB.

# Standard Diagnostics {#sec-standard}

## Stock trajectory

The full state-space trajectory with 95% confidence intervals. Recruitment, SSB, and
mean F are estimated jointly with their process uncertainty.

```{r fig-trajectory}
#| fig-cap: "Model-estimated SSB (top), mean F (middle), and recruitment (bottom) with 95% confidence intervals."
show_diag("h1_stock_trajectory.png")
```

```{r fig-trajectory-zoom}
#| fig-cap: "Recent (post-2002) stock trajectory. Shaded band = 95% CI. Dashed red line (SSB panel) = JJM h1_1.14 SSB ± 95% CI."
show_diag("h1_stock_trajectory_zoom.png")
```

## Observation variances {#sec-obsvars}

SAM estimates observation variances internally. Small values indicate a data source is
given high weight; large values indicate the model discounts that source. All catch datasets are given high weight with the offshore trawl receiving highest weight. The ratio of observation variance to coefficient of variation (CV) reflects precision versus
consistency. Generally, one doesn't want to observe low observation variance with high CVs and vice versa, indicating inappropriate model fit. 

```{r fig-obsvar}
#| fig-cap: "Internally estimated observation variances by data source, sorted by magnitude."
show_diag("h1_observation_var.png")
```

```{r fig-obsvarcv}
#| fig-cap: "Observation variance versus estimation CV. Points far to the right are down-weighted by the model."
show_diag("h1_variance_vs_cv.png")
```

## F variance parameters {#sec-fvars}

Age-coupled F parameters control how smoothly selectivity varies across ages within
each fleet. Panels show the estimated variance for each age group. In general, the best fitting model showed very similar variance across all ages. The South Central fleet showed lowest variability in F over time while the Offshore fleet showed highest variability. 

```{r fig-fvars}
#| fig-cap: "F variance parameters by fleet and age group."
show_diag("h1_fvars.png")
```

## Selectivity {#sec-sel}

Selectivity is expressed as F/Fbar at each age and fleet, grouped by five-year periods
to show temporal trends. Changes in shape over time may indicate non-stationarity in
fleet behaviour or stock demography. There is substantial change in all fleets from relatively dome shaped to increasing and vice versa. The model estimates very high selection at oldest ages in the South Central fleet which is not expected to be driven by high mortality but probably due to a lack of availability. 

```{r fig-sel}
#| fig-cap: "Estimated selectivity (F/Fbar) by fleet and five-year period. Each line is one year; colour scale = year."
show_diag("h1_selectivity.png")
```

## Fishing mortality at age {#sec-fatage}

Fleet-specific F at age since 2000. Each panel corresponds to one of the four catch
fleets.

```{r fig-fatage}
#| fig-cap: "F at age from 2000 by catch fleet. Each line is one age group."
show_diag("h1_fatage.png")
```

## Catch residuals {#sec-catchres}

Standardised residuals of the log-normal catch observation equation. Bubble area is
proportional to the magnitude of the residual; blue = positive, red = negative.
Systematic patterns (diagonal bands = cohort, horizontal bands = age, vertical bands =
year effect) indicate model misspecification. There are substantial bands to be seen by age in the FarNorth fleet (age 6+),
and the Offshore trawl (age 8+).

```{r fig-catchres}
#| fig-cap: "Standardised catch residuals by fleet. Bubble size = |residual|; blue = positive, red = negative."
show_diag("h1_catch_residuals.png")
```

## Survey index fits {#sec-surveyfit}

Observed biomass index values (points) against model-predicted values (red line) for
each survey. Persistent bias in a subset of years suggests time-varying catchability.
There is considerably misfit for most of the indices. Especially pronounced are the misfits for Acoustic North,
The Chilean CPUE, offshore CPUE and Peruvian CPUE. 

```{r fig-surveyfit}
#| fig-cap: "Survey index fits. Points = observed; red line = model prediction."
show_diag("h1_index_fits.png")
```

## Survey residuals {#sec-surveyres}

Lollipop residual plot for survey biomass indices. A trend in residuals (e.g. positive
early years and negative late years) indicates that SAM's fixed-$q$ assumption is
violated for that survey.

```{r fig-surveyres}
#| fig-cap: "Standardised survey residuals by index."
show_diag("h1_survey_residuals.png")
```

## Process error {#sec-procerr}

The process-error plots show the estimated deviation in log-numbers (top) and
log-mortality (bottom) by age and year. Large process errors indicate years where the
model needed to absorb information that the deterministic population projection could
not explain. From age 6+ there are substantial temporal patterns with age 6 & 7 being opposite in sign from age 8 & 9 which
may reflect issues with age reading. There is a general tendency to overestimate older ages. 

```{r fig-procerr-n}
#| fig-cap: "Process errors in log-N by age and year (relative, stock weight-scaled)."
show_diag("h1_procerr_N.png")
```

```{r fig-procerr-m}
#| fig-cap: "Process errors in log-mortality by age and year."
show_diag("h1_procerr_M.png")
```

## Correlation matrix {#sec-cor}

The parameter correlation matrix provides a summary of structural collinearity in the
model. There is high correlation between the between-age correlation (itrans_rho) in the F states and the variance parameters (logSdLogFsta)
as s well as within the variance parameters. All other correlations are low. 

```{r fig-cor}
#| fig-cap: "Pairwise correlation matrix of estimated parameters."
show_diag("h1_cor_params.png")
```

## Stock-recruit relationship {#sec-sr}

The SSB–recruitment scatter provides a visual check on plausible stock dynamics. The
absence of a strong Beverton-Holt or Ricker relationship is expected for short-lived
pelagic species where recruitment is highly variable. Points coloured by year; path
shows chronological sequence.

```{r fig-sr}
#| fig-cap: "Estimated SSB vs age-1 recruitment. Points coloured by year; path shows chronological sequence."
show_diag("h1_stock_recruit.png")
```

## Numbers at age {#sec-natage}

Model-estimated stock numbers at age on a log scale since 2000. Parallel lines
indicate stable age structure; divergence highlights cohort events or shifts in F.

```{r fig-natage}
#| fig-cap: "Estimated stock numbers at age (log-scale) from 2000."
show_diag("h1_natage.png")
```

# Retrospective Analysis {#sec-retro}

Retrospective analyses sequentially remove the most recent year of data and re-fit the
model. A systematic upward or downward revision of SSB or Fbar in the terminal year
indicates a retrospective pattern, quantified by Mohn's $\rho$ [@mohn1999retrospective].
In general, retrospective patterns are reasonable for SSB (which is estimated with large uncertainty) and F. 

$$
\rho = \frac{1}{n}\sum_{p=1}^{n}\frac{\hat{\theta}_{T-p,T-p} - \hat{\theta}_{T-p,T}}{\hat{\theta}_{T-p,T}}
$$

where $\hat{\theta}_{T-p,T-p}$ is the terminal-year estimate in the peel ending at year
$T-p$ and $\hat{\theta}_{T-p,T}$ is the full-model estimate for the same year. Values
of $|\rho| > 0.20$ for SSB or $|\rho| > 0.30$ for F are commonly considered
problematic [@mohn1999retrospective].

```{r fig-retro-ssb}
#| fig-cap: "Retrospective SSB trajectories (10 peels). Black line and shaded band = full model; coloured lines = retrospective peels; dashed green = JJM h1_1.14 SSB ± 95% CI."
show_diag("h1_retro_ssb.png")
```

```{r fig-retro-fbar}
#| fig-cap: "Retrospective Fbar trajectories."
show_diag("h1_retro_fbar.png")
```

```{r retro-rho, results='asis'}
rho_file <- file.path(diag_dir, "h1_mohns_rho.txt")
if (file.exists(rho_file)) {
  cat("```\n")
  cat(readLines(rho_file), sep = "\n")
  cat("\n```\n")
}
```

## In-depth retrospective diagnostics {#sec-retro-indepth}

Large retrospective patterns can arise from several sources: (i) re-estimation of catchability and observation-variance parameters as data are added; (ii) high-influence observations that cause the model to re-interpret earlier years once the terminal year is included; or (iii) surveys whose residual structure changes systematically as more data are appended. The following four diagnostics, run with residuals enabled in each retrospective peel, help identify which mechanism is dominant.

### Catchability across peels

Catchability ($\log q$) is re-estimated in each peeled model. A stable $\log q$ across peels indicates that the parameter is identified by the bulk of the time-series and is not being pulled by a small number of terminal-year observations. Large drifts — particularly in the most recent peels — suggest that new data are revising the model's view of that survey's relationship to biomass, which in turn propagates into a retrospective SSB revision. Especially the acoustic North and Peruvian CPUE show consistent trends in estimated catchability. In both cases, catcahbility increases meaning that the indices 'see' more fish. Declines are visible for the Offshore CPUE and the Chilean CPUE. 

```{r fig-retro-q}
#| fig-cap: "Estimated log-catchability per survey across retrospective peels. x-axis = terminal year of each peel; a stable line indicates the parameter is well-identified throughout the series."
show_diag("h1_retro_q_params.png")
```

### Observation variance across peels

Observation-variance ($\sigma^2$) parameters govern how much weight is placed on each data source. If $\sigma^2$ for a survey drifts as years are removed, it signals that the noise structure inferred from recent data differs from the historical period — often a sign of a trend in data quality or catchability that the model absorbs into the variance estimate rather than the mean. The observation variance of the offshore and South Cental fishery has increased considerably over the past 5 years, but are still estimated at low levels. There seems to be some model instability which shows as sudden drops in observation variance for the Far North fleet. 

```{r fig-retro-obsvar}
#| fig-cap: "Estimated observation variances across retrospective peels, split by catch fleets (top) and survey indices (bottom)."
show_diag("h1_retro_obsvar_params.png")
```

### Survey residual spaghetti

Each line shows the standardised residuals of a survey for all years from one retrospective peel. The black line is the full model. If residuals in the *middle* of the time-series shift between peels, earlier data are being re-interpreted as terminal observations change — a structural conflict between that survey and others. If only the most terminal years shift, the pattern is consistent with normal retrospective behaviour driven by low information content at the data edge.

```{r fig-retro-spaghetti}
#| fig-cap: "Standardised survey residuals from all retrospective peels overlaid. Black line = full model; colour = terminal year of peel. Divergence from the full model in non-terminal years indicates structural conflict."
show_diag("h1_retro_residual_spaghetti.png")
```

```{r fig-retro-catch-spaghetti}
#| fig-cap: "Mean catch-fleet residuals (mean across ages per year) across retrospective peels."
show_diag("h1_retro_catch_residual_spaghetti.png")
```

### Residual stability heatmap

For each survey and year, the standard deviation of the standardised residual across all peels that included that year is computed. A year with high SD is re-interpreted substantially each time a new year of data is added — i.e. it has high leverage on the model trajectory. Years and surveys with consistently high SD are the primary suspects driving the retrospective pattern.

```{r fig-retro-stability}
#| fig-cap: "SD of standardised residuals across all retrospective peels, by survey and year. High values (yellow/bright) indicate high-leverage observations that shift interpretation as more data are added."
show_diag("h1_retro_residual_stability.png")
```

### Terminal-year residuals: drivers of the retrospective pattern

The retrospective pattern arises because the model mis-fits certain observations at the terminal year of each peel. The plot below shows, for each fleet and each peel, the ratio of the observed to model-predicted biomass at that peel's terminal year. A ratio consistently above 1 (blue) means the model persistently underpredicts that fleet's terminal observation — causing it to revise the stock trajectory upward when a new year is added, which produces a negative Mohn's ρ for SSB. The reverse applies for ratios below 1 (red). Surveys or catch fleets with consistently one-sided bars are the primary drivers of the retrospective.

```{r fig-retro-topchange}
#| fig-cap: "Observed / predicted biomass ratio at the terminal year of each retrospective peel, per fleet. Blue = model underpredicts; red = overpredicts. Consistent direction = systematic mis-fit driving the retrospective pattern."
#| fig-height: 8
show_diag("h1_retro_residual_topchange.png")
```

# Leave-one-out Analysis {#sec-loo}

Leave-one-out (LOO) analysis refits the model seven times, each time omitting one
survey group entirely. For the three surveys with JJM q-break years (Chile_AcousCS,
Chile_CPUE, Offshore_CPUE), both the early and late sub-series are dropped together
so that each LOO run represents a genuine removal of that data source.

The resulting SSB, Fbar, and recruitment trajectories are overlaid on the full model
estimate. A large shift when a survey is dropped indicates that the model is strongly
influenced by that data source — either because it carries substantial information or
because it is in conflict with other inputs. Surveys whose removal causes little change
are relatively redundant given the remaining data.

```{r fig-loo-ssb}
#| fig-cap: "Leave-one-out SSB. Black line = full model; coloured lines = model with that survey group dropped (no CI shown). Dashed green = JJM h1_1.14 SSB."
show_diag("h1_loo_ssb.png")
```

```{r fig-loo-fbar}
#| fig-cap: "Leave-one-out mean F."
show_diag("h1_loo_fbar.png")
```

```{r fig-loo-rec}
#| fig-cap: "Leave-one-out recruitment."
show_diag("h1_loo_rec.png")
```

# Catchability Diagnostics {#sec-q}

Fixed catchability ($q$) is one of the most consequential assumptions in any assessment
model. If the true $q$ of a survey changes over time — due to changes in gear
efficiency, stock distribution, or survey design — the model will accumulate systematic
residuals. The following diagnostics are designed to detect and characterise such
trends. 

## Rolling residuals and catchability drift {#sec-qroll}

A five-year rolling mean of standardised residuals reveals slow shifts in effective
catchability. Persistent positive residuals indicate the model is under-predicting the
survey (i.e. $q$ is in reality increasing), and vice versa. The figure clearly shows
that several indices suffer from trends in residuals, specifically the the Chilean CPUE, offshore CPUE,
Peruvian Acoustic. 

```{r fig-qroll}
#| fig-cap: "Survey residuals (bars) with 5-year rolling mean (line). A trend in the rolling mean indicates time-varying catchability."
show_diag("h1_q_rolling_residuals.png")
```

## Catchability shift at break years {#sec-qshift}

The JJM `h1_1.14` control file assigns discrete q-break years to three indices
(Chile_AcousCS: 2002, Chile_CPUE: 2000, Offshore_CPUE: 2021). To honour these
structural breaks, each of these indices is split into an early and a late
sub-series in SAM, with the two sub-series sharing the same observation variance
parameter but receiving independent catchability estimates. The figure below shows
the estimated log(q) for the early and late periods; a large shift indicates that
the survey efficiency or stock availability changed substantially at the break year.

```{r fig-qshift}
#| fig-cap: "Estimated log-catchability for early (blue) and late (red) sub-series of the three indices with JJM q-break years. Lines connect the two periods within each index."
show_diag("h1_catchability_change.png")
```

## Implied catchability trend {#sec-qtrend}

The quantity $\log(I_t) - \log(\hat{B}_t)$ is proportional to $\log(q_t)$. Centring
this series by its mean isolates the relative change in $q$ over time. A significant
linear trend (shown in grey) indicates that the constant-$q$ assumption is violated.
Only the Chilean Acoustic CS, the DEPM and the Peruvian CPUE show no significant trend. 

```{r fig-qtrend}
#| fig-cap: "Implied log-catchability relative to fleet mean. Slope ≠ 0 indicates time-varying q. OLS trend shown in black."
show_diag("h1_q_trend.png")
```

## Runs test {#sec-runs}

The Wald-Wolfowitz runs test [@methot2013stock] evaluates whether the signs of
residuals are distributed randomly. Significantly too few runs (p < 0.05) indicates
a non-random pattern, most commonly caused by time-varying catchability or model
misspecification.

```{r fig-runs}
#| fig-cap: "Runs test p-value for each fleet. Bars below the dashed line (p = 0.05) indicate non-random residuals."
show_diag("h1_runs_test.png")
```

```{r tbl-runs, results='asis'}
runs_file <- file.path(diag_dir, "h1_runs_test.csv")
if (file.exists(runs_file)) {
  rt <- read.csv(runs_file)
  knitr::kable(rt, caption = "Runs test results by fleet.", digits = 4)
}
```

# Fleet Conflict and Compensation Diagnostics {#sec-conflict}

When multiple fleets and surveys contribute information to the same population
trajectory, they may provide conflicting signals. SAM's internal weighting resolves
conflicts by down-weighting inconsistent sources, but residual patterns can still
reveal the nature of the tension.

## Residual cross-correlation matrix {#sec-crosscor}

The cross-correlation of annual (mean) residuals between all fleets identifies
structural conflicts. A strong negative correlation between two fleets means one
over-predicts systematically when the other under-predicts — a diagnostic signal for
fleet compensation. Positive correlations suggest fleets respond coherently to the
same environmental or stock signal. Especially the offshore CPUE contradicts in trend with 
the Chilean SC catch data, although the correlation with the Chilean SC CPUE is high. 

```{r fig-crosscor}
#| fig-cap: "Cross-correlation of mean annual residuals between all fleets. Red = negative correlation (conflict); blue = positive (co-vary)."
show_diag("h1_residual_crosscor.png")
```

## Pairwise survey index scatter {#sec-pairwise}

Plotting pairs of survey indices over shared years shows whether surveys agree on the
direction and magnitude of stock change. Diverging trends or low correlations indicate
that surveys sample different components of the stock or are affected by different
sources of variability.

```{r fig-pairwise}
#| fig-cap: "Pairwise scatter of log-biomass survey indices. Points coloured by year; OLS trend shown."
show_diag("h1_survey_pairwise.png")
```

## Cohort tracking across catch fleets {#sec-cohort}

The same cohort should be detectable across all catch fleets as it ages through the
population. If catch-at-age data from different fleets show diverging relative
abundance for the same cohort at the same age, it indicates either gear selectivity
conflicts, spatial overlap differences, or data inconsistencies. There is very poor cohort tracking capabilities for the different catch@age data. For example, the 2018 cohort is picked up markedly different by 3 out of the 4 fleets. 

```{r fig-cohort}
#| fig-cap: "Relative catch-at-age by fleet for recent cohorts. Each fleet is a colour; divergence between lines indicates fleet conflict on cohort strength or selectivity."
show_diag("h1_cohort_tracking.png")
```

# Productivity Analysis {#sec-productivity}

## Rationale

The Beverton-Holt stock-recruitment relationship describes how spawning biomass (SSB)
translates into future recruitment. The steepness parameter $h$ is the fraction of
unfished recruitment $R_0$ that is produced at 20% of unfished SSB ($B_0$):

$$
R = \frac{0.8\,R_0\,h\,\text{SSB}}{0.2\,B_0(1-h) + (5h-1)\,\text{SSB}}
\quad\equiv\quad
R = \frac{a\,\text{SSB}}{b + \text{SSB}},
\quad h = \frac{0.2(b + B_0)}{b + 0.2\,B_0}
$$

where $B_0$ is approximated by the maximum observed SSB. High steepness ($h > 0.8$)
indicates low compensation — recruitment is nearly independent of spawning biomass.

A change in steepness over time signals a **shift in stock productivity** that is not
captured by the primary assessment model (JJM uses fixed life-history parameters). Such
shifts can arise from regime changes, density-dependent effects at low stock size, or
environmental shifts affecting larval survival.

## Method

Beverton-Holt curves are fitted by **weighted nonlinear least squares**, using
uncertainty in the SAM recruitment estimate as observation weights:

$$
w_t = \frac{1}{\hat{\sigma}^2_{\log R_t}}, \qquad
\hat{\sigma}_{\log R_t} \approx \frac{\text{CI}_{97.5} - \text{CI}_{2.5}}{4\,R_t}
$$

Change points are found by exhaustive search over all candidate break years (minimum
10 years per segment). The improvement in total weighted residual SS when adding a
break is tested with an $F$-test (2 additional parameters per break). A second break
is only searched if the first is significant at $p < 0.05$.

```{r prod-summary, results='asis'}
sf <- file.path(diag_dir, "h1_productivity_summary.txt")
if (file.exists(sf)) {
  cat("```\n")
  cat(readLines(sf), sep = "\n")
  cat("\n```\n")
}
```

## Productivity over time

Annual productivity (R/SSB) with 95% confidence intervals propagated from SAM
uncertainty in both recruitment and SSB via the delta method on the log scale:
$\text{SE}_{\log(R/\text{SSB})} \approx \sqrt{\text{SE}^2_{\log R} + \text{SE}^2_{\log \text{SSB}}}$.
A LOESS smoother highlights the long-term trend without imposing a parametric form.

```{r fig-prod-time}
#| fig-cap: "Productivity (R/SSB) over time. Points = annual estimate; bars = 95% CI; line = LOESS smoother. A declining trend indicates reducing stock productivity."
show_diag("h1_productivity_time.png")
```

## Stock-recruit panels

Up to three panels are shown depending on how many break points are statistically
supported. The legend of each multi-period panel includes the estimated steepness for
that period. Error bars show 95% confidence intervals from SAM on both SSB (horizontal)
and recruitment (vertical).

```{r fig-productivity}
#| fig-cap: "Beverton-Holt stock-recruit analysis. Panel 1: full time series (points coloured by year). Panel 2 (if shown): best single break, two B-H curves coloured by period with steepness in legend. Panel 3 (if shown): two breaks, three periods."
show_diag("h1_productivity_sr.png")
```

# Selectivity Block Suggestions for JJM {#sec-blocks}

## Rationale

JJM requires the user to pre-specify constant-selectivity blocks: periods within which
selectivity-at-age is assumed to be time-invariant. In practice these blocks are set
based on expert knowledge of fleet changes, but there is no objective statistical
criterion for where boundaries should fall.

SAM estimates selectivity annually through the random-walk F process. The trajectory
of the annual selectivity *shape* (F/Fbar at age, normalised so the peak = 1) therefore
provides an empirical signal of when selectivity changed significantly. A structural
break in this trajectory is a principled suggestion for a new JJM block boundary.

## Method

For each catch fleet:

1. The normalised selectivity-at-age vector is extracted for every year from the SAM
   fit (shape only: F/Fbar at age, scaled so the peak age = 1).
2. The dominant mode of temporal variation is captured by the first principal component
   (PC1) of the annual selectivity matrix.
3. The PC1 time series is split into blocks using **binary segmentation**, with each
   year weighted by its relative catch ($w_t = C_t / \max C$). This ensures that
   periods with little fishing activity have minimal influence on where block
   boundaries are placed. Splits require $p < 0.05$ (F-test) and at least five years
   per segment.
4. Each detected breakpoint is assigned an **importance score**:

$$
\text{importance} = F_{\text{split}} \times \frac{\sum_{t \in \text{segment}} C_t}{C_{\text{total}}}
$$

   Breakpoints are **ranked 1 (most important) downward**: rank 1 should be kept even
   if the number of blocks is reduced; rank $k$ can be dropped first. This allows a
   user to move from the full block structure to a simpler one by dropping the
   highest-numbered ranks.
5. Within-block variance, peak-selectivity age, and the percentage of total fleet catch
   in each block are reported as supporting statistics.

## PC1 trajectories and block boundaries

The PC1 score over time reveals the dominant mode of selectivity change. Vertical red
lines mark suggested block boundaries; the **number is the importance rank** (1 = most
important). Grey shading behind the line shows the relative catch weight. To reduce the
number of blocks, drop boundaries starting from the highest rank number.

```{r fig-sel-pc1}
#| fig-cap: "PC1 of annual selectivity shape per fleet. Red lines = block boundaries; rank label (1 = most important to retain). Grey shading = relative catch weight."
show_diag("h1_selectivity_pc1.png")
```

## Selectivity curves by block

Individual annual selectivity curves are shown in the colour of their assigned block;
the dashed line is the block mean. Blocks with tightly clustered curves support the
constant-selectivity assumption; blocks where thin lines diverge strongly from the mean
suggest that within-block variation may still be substantial.

```{r fig-sel-byblock}
#| fig-cap: "Annual selectivity curves coloured by suggested JJM block. Dashed line = block mean."
show_diag("h1_selectivity_byblock.png")
```

## Suggested block table

```{r tbl-blocks, results='asis'}
blk_file <- file.path(diag_dir, "h1_selectivity_blocks.csv")
if (file.exists(blk_file)) {
  blk <- read.csv(blk_file)
  # Last block of each fleet has no trailing boundary → rank is NA; show as "—"
  blk$boundary_rank <- ifelse(is.na(blk$boundary_rank), "\u2014",
                               as.character(blk$boundary_rank))
  knitr::kable(
    blk[, c("fleet", "block", "start_year", "end_year",
            "n_years", "catch_pct", "peak_age", "within_var",
            "boundary_rank", "rationale")],
    col.names = c("Fleet", "Block", "Start", "End", "N years",
                  "Catch %", "Peak age", "Within-var.",
                  "Boundary rank", "Rationale"),
    caption   = paste(
      "Suggested constant-selectivity blocks for JJM by catch fleet.",
      "**Catch %** = percentage of total fleet catch in that block.",
      "**Boundary rank** = importance of the boundary ending this block",
      "(1 = most important, highest rank = drop first to reduce blocks).",
      "'—' = final block (no subsequent boundary).",
      "Within-block variance is the mean temporal variance in normalised",
      "selectivity across ages; lower = more stable selectivity shape."),
    digits    = 5
  )
}
```

## Caveats

- The analysis operates on SAM's estimated selectivity, which itself carries
  uncertainty. Short blocks (< 5 years) or blocks separated by a weak PC1 shift should
  be treated with caution.
- SAM's random-walk F tends to smooth selectivity; abrupt one-year anomalies visible
  in the raw data may be absorbed into the process error rather than reflected in the
  selectivity trajectory.
- The suggestions are a starting point. Fleet-specific information (gear changes, area
  closures, quota redistribution) should take precedence when it conflicts with the
  statistical signal.

# Discussion {#sec-discussion}

## Utility of SAM as a diagnostic tool

The primary value of applying SAM alongside JJM is not to replace the primary
assessment but to leverage SAM's internal observation-variance estimation to identify
data-weighting inconsistencies. Where SAM assigns a high observation variance to a
particular fleet, it suggests that the fleet's contribution to the likelihood is
relatively small — a finding that may inform external weighting decisions in JJM.

The process-error plots (Section @sec-standard) are particularly informative: large
process errors in specific age classes or years indicate that the deterministic
population projection is insufficient to track the observed data, which may signal
recruitment events, migration, or measurement issues not captured in the model
structure.

## Interpreting catchability diagnostics

The implied-$q$ diagnostics (Section @sec-q) should be interpreted with care for
biomass-only surveys covering large and spatially heterogeneous stock distributions.
The ratio $\log(I_t) - \log(\hat{B}_t)$ conflates true catchability change with
changes in stock distribution, survey coverage, and gear efficiency. A statistically
significant trend in this quantity is a necessary but not sufficient condition for
concluding that survey $q$ has changed.

For the runs test, non-random residuals (p < 0.05) are common in long biomass survey
series covering stocks with high interannual variability, particularly when the fixed-$q$
model cannot fully describe index dynamics. These results should be treated as prompts
for investigation rather than definitive evidence of model failure.

## Fleet conflict and compensation

Negative entries in the residual cross-correlation matrix (@sec-crosscor) indicate
that two fleets pull the model in opposite directions. If stock distribution shifts northward
in some years, one survey will over-index while another under-indexes, even if total
abundance is unchanged.

SAM resolves these conflicts through differential down-weighting, but the resolution
implies that the "losing" fleet contributes less to the final trajectory estimate.
Pairwise scatter plots (@sec-pairwise) and cohort tracking (@sec-cohort) help identify
which surveys are internally consistent with each other and which represent independent
or contradictory signals.

## Limitations of the SAM diagnostic

Several aspects of the JJM assessment cannot be directly evaluated through SAM:

- **Length composition data** are not used in the SAM fit. JJM uses both age and length
  compositions, and the interaction between the two likelihood components is an
  important source of information on the growth transition matrix.
- **Biomass-index simplification**: JJM surveys may be age-disaggregated or
  vulnerability-weighted, whereas the SAM diagnostic treats all indices as
  integrated biomass. This may absorb selectivity changes into the $q$ term.
- **Stock-recruitment**: SAM does not impose a stock-recruitment relationship; JJM can
  optionally constrain recruitment. The SAM retrospective therefore reflects pure
  data-driven retrospective bias rather than the interplay between SR assumptions and
  data.
- **Two-stock configuration (H2)**: This analysis covers H1 only. A separate SAM run
  for H2's far-northern stock would be needed to evaluate the split-stock hypothesis.

## Recommendations

Based on the diagnostic results, the following assessments are suggested:

1. **Observation variance comparison**: Compare the relative magnitudes of SAM's
   estimated observation variances to the effective sample sizes used in JJM.
   Systematic disagreement may indicate that JJM over- or under-weights certain
   surveys.
2. **Catchability trend investigation**: For any survey flagging a significant runs-test
   failure or a linear trend in implied $q$, review survey design changes, spatial
   coverage, and equipment upgrades over the time series.
3. **Fleet conflict investigation**: Pairs of surveys with strongly negative
   cross-correlations in residuals should be examined for spatial stratification
   differences and temporal coverage gaps.
4. **Retrospective pattern**: If Mohn's $\rho$ for SSB exceeds 0.20 in absolute value,
   investigate the most recent data years for influential observations or catchability
   breaks.

# References {.unnumbered}
